Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-02-12 Multiresolution_framework/ltx/manuscript.tex
some re-arrangement in EM

For the linear state-space model given by Eqs.~\eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}, the bound defined by Eq.~\eqref{eq:Bishopbound} is a quadratic of the form (see \ref{ap:QDerivation} for derivation) 
\begin{equation}\label{eq:MRA-QCompact}
\mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\varsigma\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\varsigma\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
where $\beta$ is constant with respect to $\boldsymbol\theta$. Equation~\eqref{eq:MRA-QCompact} is maximum at
\begin{align}\label{eq:MRA-thetahat}
\boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1)^\top,
\end{align}
where
\begin{align}
\boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\boldsymbol\Gamma_1]_{j,i}\label{eq:epsilon0} \\
\boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\boldsymbol\Gamma_1]^{j,i} \label{eq:epsilon1}\\
\boldsymbol\Upsilon&=T_s\varsigma\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\boldsymbol\Gamma_2]^{j,i},\label{eq:Epsilon}
\end{align} 
where $[\cdot]_{p,q}$ denotes the $\left(p,q\right)$-element of the matrix, and $ [\cdot]^{p,q}$ denotes the $\left(p,q\right)$-block of the block matrix and where 
\begin{align}
\left[ \Gamma_1\right]^{j,i} &=\sum_{k=1}^{n_x}\left[ \boldsymbol\Sigma_w^{-1}\boldsymbol\Lambda_x^{-1}\right]_{jk} \left[ \mathbf U\right]^{k,i},\\
\left[ \Gamma_2\right] ^{j,i}&=\sum_{k,m=1}^{n_x}[\mathbf U^{\top}]^{jk} \left[\boldsymbol\Lambda_x^{-1}\boldsymbol\Sigma_w^{-1}\boldsymbol\Lambda_x^{-1} \right]_{km}[\mathbf U]^{mi}.
\end{align}
Each $ 1 \times n_{\theta}$ block of the $n_x \times n_x n_{\theta}$ block matrix $\mathbf U$ is 
\begin{align}\label{eq:Uij}
\left[ \mathbf U\right] ^{k,i}&=\int_{\boldsymbol \Omega}\left[\boldsymbol\mu(r) \right]_k \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(r-r'\right) \mathrm{d}r'\right]_{i:} \mathrm{d}r,
\end{align}
where $[\cdot]_{i:} $ denotes the $i^{th}$ row. The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ in Eqs.~\eqref{eq:epsilon0}-\eqref{eq:Epsilon} are calculated using the RTS smoother outputs: $\hat{\mathbf x}_t$, covariance, $\mathbf P_t^b$, and cross-covariance matrix, $\mathbf M_t$ \citep{Gibsona2005}, 
\begin{align}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:Xi0} \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).  \label{eq:Xi1}
\end{align}

\subsection{Stopping rule}
The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \citep{McLachlan1997}. The stopping rule adopted herein is
\begin{equation}
 \left(\parallel \mathbf{A} \parallel_{F}^{(i)}-\parallel \mathbf{A} \parallel_{F}^{(i-1)}\right)<\epsilon,
 \end{equation}
 where $\epsilon$ is a threshold value and $\parallel \mathbf{A} \parallel_{F}^{(i)}$ and $ \parallel \mathbf{A} \parallel_{F}^{(i-1)}$ are the Frobenius norms  of the successive estimates of $\mathbf{A} $ matrices which is defined by \citet{Meyer2000}
 \begin{equation}
  \parallel \mathbf{A} \parallel_{F}=\sqrt{\sum_{i,j=1}^{n_x}\mid a_{i,j} \mid^2}=\sqrt{\mathrm{tr} (\mathbf A^{\top}\mathbf A)}.
 \end{equation}
\subsection{Computational complexity} 
All $n_x^2$ blocks of $\mathbf U$, $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$  can be computed as a one-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the implementation of \citet{Dewar2009}. The complexity of each step of the algorithm is summarized in Table~\ref{table:MRA-ComputationalComplexity} where for ease of comparison the computational complexity of the method of \citet{Dewar2009} is also identified. It can be seen from the table that the state dimension is an important factor at each iteration of the estimation algorithm, emphasizing the advantage of the proposed algorithm compared to the method proposed in \citet{Dewar2009}. This also suggests the use of an efficient algorithm to choose the minimum number of basis functions (states of the system) while achieving the desired details of the underlying field.  
\begin {table}
\begin{center}
\scalebox{1}{\begin{tabular}{lllc}
\hline \hline
Variable &Equation&Order&Order obtained from \citet{Dewar2009}\\ \hline\\
$\Xi_0, \Xi_1$&Eqs.~\eqref{eq:Xi0} and \eqref{eq:Xi1}&$O(Tn_x^2)$& \----\\
$\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$&Eqs.~\eqref{eq:epsilon0} and \eqref{eq:epsilon1} &$O(n_x^2n_{\theta})$ &$O(n_x^2n_{\theta})+O(n_x^3)$ \\
$\boldsymbol\Upsilon$&Eq.~\eqref{eq:Epsilon}&$O(n_x^2n_{\theta}^2)$&$O(n_x^4n_{\theta}^2)$\\
$\mathbf A(\boldsymbol\theta)$&Eq.~\eqref{eq:A_theta}&$O(n_x^3)$&\----\\
$\boldsymbol\Lambda_{\theta}$&Eq.~\eqref{eq:LambdaThetaElements}&$O(n_x^2n_{\theta})$&\----\\
$\hat{\boldsymbol\theta} $&Eq.~\eqref{eq:MRA-thetahat}&$O(n_{\theta}^3)$&\----\\
\hline \hline
\end{tabular}}
 \caption {{\bf The Computational complexity of the estimation algorithm}. A comparison between the algorithm proposed herein and that of \citet{Dewar2009} is also provided. "\----'' indicate equal complexities for both algorithms.} 
\label{table:MRA-ComputationalComplexity}
\end{center}
\end {table}

@@

>>>>

For the linear state-space model given by Eqs.~\eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}, the bound defined by Eq.~\eqref{eq:Bishopbound} is a quadratic of the form (see \ref{ap:QDerivation} for derivation) 
\begin{equation}\label{eq:MRA-QCompact}
where $\beta$ is constant with respect to $\boldsymbol\theta$. Equation~\eqref{eq:MRA-QCompact} is maximum at
\begin{align}\label{eq:MRA-thetahat}
\boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1)^\top,
\boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\boldsymbol\Gamma_1]_{j,i}\label{eq:epsilon0} \\
\boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\boldsymbol\Gamma_1]^{j,i} \label{eq:epsilon1}\\
\boldsymbol\Upsilon&=T_s\varsigma\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\boldsymbol\Gamma_2]^{j,i},\label{eq:Epsilon}
where $[\cdot]_{p,q}$ denotes the $\left(p,q\right)$-element of the matrix, and $ [\cdot]^{p,q}$ denotes the $\left(p,q\right)$-block of the block matrix and where 
\left[ \Gamma_1\right]^{j,i} &=\sum_{k=1}^{n_x}\left[ \boldsymbol\Sigma_w^{-1}\boldsymbol\Lambda_x^{-1}\right]_{jk} \left[ \mathbf U\right]^{k,i},\\
\left[ \Gamma_2\right] ^{j,i}&=\sum_{k,m=1}^{n_x}[\mathbf U^{\top}]^{jk} \left[\boldsymbol\Lambda_x^{-1}\boldsymbol\Sigma_w^{-1}\boldsymbol\Lambda_x^{-1} \right]_{km}[\mathbf U]^{mi}.
Each $ 1 \times n_{\theta}$ block of the $n_x \times n_x n_{\theta}$ block matrix $\mathbf U$ is 
\begin{align}\label{eq:Uij}
\left[ \mathbf U\right] ^{k,i}&=\int_{\boldsymbol \Omega}\left[\boldsymbol\mu(r) \right]_k \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(r-r'\right) \mathrm{d}r'\right]_{i:} \mathrm{d}r,
where $[\cdot]_{i:} $ denotes the $i^{th}$ row. The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ in Eqs.~\eqref{eq:epsilon0}-\eqref{eq:Epsilon} are calculated using the RTS smoother outputs: $\hat{\mathbf x}_t$, covariance, $\mathbf P_t^b$, and cross-covariance matrix, $\mathbf M_t$ \citep{Gibsona2005}, 
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:Xi0} \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).  \label{eq:Xi1}

@@

\subsection{Stopping rule}
The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \citep{McLachlan1997}. The stopping rule adopted herein is
 \left(\parallel \mathbf{A} \parallel_{F}^{(i)}-\parallel \mathbf{A} \parallel_{F}^{(i-1)}\right)<\epsilon,
 where $\epsilon$ is a threshold value and $\parallel \mathbf{A} \parallel_{F}^{(i)}$ and $ \parallel \mathbf{A} \parallel_{F}^{(i-1)}$ are the Frobenius norms  of the successive estimates of $\mathbf{A} $ matrices which is defined by \citet{Meyer2000}
  \parallel \mathbf{A} \parallel_{F}=\sqrt{\sum_{i,j=1}^{n_x}\mid a_{i,j} \mid^2}=\sqrt{\mathrm{tr} (\mathbf A^{\top}\mathbf A)}.
\subsection{Computational complexity} 
All $n_x^2$ blocks of $\mathbf U$, $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$  can be computed as a one-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the implementation of \citet{Dewar2009}. The complexity of each step of the algorithm is summarized in Table~\ref{table:MRA-ComputationalComplexity} where for ease of comparison the computational complexity of the method of \citet{Dewar2009} is also identified. It can be seen from the table that the state dimension is an important factor at each iteration of the estimation algorithm, emphasizing the advantage of the proposed algorithm compared to the method proposed in \citet{Dewar2009}. This also suggests the use of an efficient algorithm to choose the minimum number of basis functions (states of the system) while achieving the desired details of the underlying field.  
\begin {table}
\scalebox{1}{\begin{tabular}{lllc}
Variable &Equation&Order&Order obtained from \citet{Dewar2009}\\ \hline\\
$\Xi_0, \Xi_1$&Eqs.~\eqref{eq:Xi0} and \eqref{eq:Xi1}&$O(Tn_x^2)$& \----\\
$\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$&Eqs.~\eqref{eq:epsilon0} and \eqref{eq:epsilon1} &$O(n_x^2n_{\theta})$ &$O(n_x^2n_{\theta})+O(n_x^3)$ \\
$\boldsymbol\Upsilon$&Eq.~\eqref{eq:Epsilon}&$O(n_x^2n_{\theta}^2)$&$O(n_x^4n_{\theta}^2)$\\
$\mathbf A(\boldsymbol\theta)$&Eq.~\eqref{eq:A_theta}&$O(n_x^3)$&\----\\
$\boldsymbol\Lambda_{\theta}$&Eq.~\eqref{eq:LambdaThetaElements}&$O(n_x^2n_{\theta})$&\----\\
$\hat{\boldsymbol\theta} $&Eq.~\eqref{eq:MRA-thetahat}&$O(n_{\theta}^3)$&\----\\
\end{tabular}}
 \caption {{\bf The Computational complexity of the estimation algorithm}. A comparison between the algorithm proposed herein and that of \citet{Dewar2009} is also provided. "\----'' indicate equal complexities for both algorithms.} 
\label{table:MRA-ComputationalComplexity}
